### 2025-05-21 docs/article/Sections/state-of-the-art.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)
chapter 2 - resuming table with a characteristics comparison among Java template engines

\textit{Web templates} have been the most widely adopted approach for
constructing dynamic HTML pages.
Web templates (such as JSP, Handlebars, or Thymeleaf), also known as
\textit{web views}~\cite{Fowler02,Alur01}, are based on
HTML documents augmented with template-specific markers (e.g., \texttt{<\%>},
\texttt{\{\{\}\}}, or \texttt{\$\{\}}), which represent \textit{dynamic}
information to be replaced at runtime with the results of corresponding
computations, producing the final HTML page.
The process of parsing and replacing these markers---i.e.,
\textit{resolution}---is the primary responsibility of the \textit{template
engine}~\cite{Parr04}.

One key characteristic of \textit{web templates} is their ability to receive a 
\textit{context object}---equivalent to the \textit{model} in the model-view 
design pattern~\cite{mvc88,Parr04}---which provides the data used to fill 
template placeholders at runtime.

Web templates can be distinguished by several properties, namely:
\begin{enumerate}
    \item Templating idiom
    \item Supported data model APIs
    \item Asynchronous support
    \item Type safety and HTML safety
    \item Progressive rendering
\end{enumerate}

Although some of the aforementioned characteristics apply to both server-side 
and client-side approaches, we focus solely on web template technologies for 
server-side rendering, as our work is centered on that approach.

Before detailing each of the aforementioned characteristics, Table~11 presents 
a breakdown of state-of-the-art template engines, classified according to the 
identified properties.

>>>>

\textit{Web templates} have been the most widely adopted approach for
constructing dynamic HTML pages.
Web templates or \textit{web views}~\cite{Fowler02,Alur01} (e.g. JSP, Handlebars,
or Thymeleaf), are based on HTML documents augmented with template-specific
markers (e.g., \texttt{<\%>}, \texttt{\{\{\}\}}, or \texttt{\$\{\}}), which
represent \textit{dynamic} information to be replaced at runtime with the
results of corresponding computations, producing the final HTML page.
The process of parsing and replacing these markers---i.e.,
\textit{resolution}---is the primary responsibility of the \textit{template
engine}~\cite{Parr04}.
One key characteristic of \textit{web templates} is their ability to receive a 
\textit{context object}---equivalent to the \textit{model} in the model-view 
design pattern~\cite{mvc88,Parr04}---which provides the data used to fill 
template placeholders at runtime.
Web templates can be distinguished by several properties, namely:
\begin{enumerate}
    \item \textit{Domain-specific language} idiom
    \item Supported data model APIs
    \item Asynchronous support
    \item Type safety and HTML safety
    \item Progressive rendering
\end{enumerate}

Although some of the aforementioned characteristics apply to both \textit{server-side}
and \textit{client-side} approaches, we focus solely on web template technologies for
\textit{server-side rendering}, as our work is centered on that approach.
Before getting deep on each of the aforementioned characteristics,
Table~\ref{table:cmplibs} presents a breakdown of mainstream template engines,
classified according to the identified properties.

\begin{table}[h]
  \small
	\tabcolsep=0.1cm
	\def\arraystretch{1.2}
	\begin{tabular}{|c|c|c|c|c|c|c|}
		\hline                                                                  
		\textbf{Library}
		&\textbf{DSL idiom}
		&\textbf{Data Model APIs}
    &\shortstack{\textbf{Asynchronous}\\\textbf{Support}}
    &\shortstack{\textbf{Type}\\\textbf{Safety}}
    &\shortstack{\textbf{HTML}\\\textbf{Safety}}
		&\shortstack{\textbf{Progressive}\\\textbf{Rendering}}
		\\
		\hline
		\textbf{Freemarker}
		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
		\textbf{JSP}
		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{red}{\times}$}
    \\
    \hline
		\textbf{JStachio}
		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\\hline
		\textbf{Pebble}
		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
		\textbf{Rocker}
		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
		\textbf{Thymeleaf}
		&External DSL
		&\shortstack{\texttt{Iterable}\\\texttt{Publisher}}
		&\texttt{Publisher}$^{(1)}$
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
		\textbf{Trimou}
		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
		\textbf{Velocity}
		&External DSL
		&\texttt{Iterable}
		&\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
    \shortstack{\textbf{Groovy }\\\textbf{MarkupBuilder}}
		&Nested Lazy
		&All
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{red}{\times}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
    \textbf{HtmlFlow}
		&Chaining
		&All
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
    \hline
		\textbf{j2html}
		&Nested Eager
		&All
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{red}{\times}$}
		\\
    \hline
		\textbf{KotlinX}
		&Nested Lazy
		&All
    &\large{$\textcolor{red}{\times}$}
    &\large{$\textcolor{PineGreen}{\checkmark}$}
    &\textcolor{PineGreen}{\checkmark}$^{(2)}$
		&\large{$\textcolor{PineGreen}{\checkmark}$}
		\\
		\hline                                                      
	\end{tabular}
	\caption{
    Comparison of web template technologies in the Java ecosystem.
    \\$^{(1)}$ Limited to a single model per web template and compatible with
    the reactive stream \texttt{Publisher}.
    \\$^{(2)}$ Non-safety for HTML attributes.
  }
	\label{table:cmplibs}
\end{table}

The first half of Table~\ref{table:cmplibs} lists template engines that use their
own templating dialects, referred to as \textit{external DSLs}. The second half
lists Java libraries that provide an \textit{internal DSL}, typically using a
\textit{nested} or \textit{chaining} style to build HTML.
Note that, by leveraging the host language (e.g., Groovy, Java, or Kotlin) as
the templating idiom, the latter impose no restrictions on the data model and
fully support all styles of data access APIs.
